---
title: "MY457/MY557: Causal Inference for Experimental and Observational Studies"
subtitle: "Seminar 5: Difference-in-Differences"
author: ""
date: ''
output:
  
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The coding exercise this week will walk through some of the core ideas of difference-in-differences (DID). First, we will load in some required packages and set a seed for reproducibility.

```{r setup,message=FALSE,warning=FALSE}
package_check <- function(need = c()){
  have <- need %in% rownames(installed.packages()) # checks packages you have
  if(any(!have)) install.packages(need[!have]) # install missing packages
  invisible(lapply(need, library, character.only=T))
}

package_check(c("dplyr", "ggplot2", "estimatr", "fixest", "plm", "panelView", "did", "fect"))

# dplyr for data management
# ggplot2 for graphing
# estimatr some nice functions for estimating treatment effects in experimental settings
# fixest for fast fixed effects
# plm, an alternative to fixest
# panelView for visualising panel data
# did gives us callaway-sant'anna estimator
# fect for counterfactual estimators
set.seed(02139)
```

# Canonical 2x2 Difference-in-Differences

## Introduction

Assume we have two time periods ($T = 0, 1$) and two groups ($G = 0, 1$). We are interested in the effect of a treatment $Z$, which switches on in time period $T=1$, on an outcome $Y$. We will assume a panel data setting in which units are observed twice.

## Simulating Data

Given this setup, we can now simulate a dataset with 1000 observations. We will first simulate the potential outcomes $Y$ as a function of $T$, $Z$, and an error term.

```{r simulate_data_2x2}
N = 1000
ATE = 2

data = tibble(
  # First, create a unit id such that each observation is observed three times:
  id = sort(rep(1:N, 3)),
  # Now, generate our time variable such that each observation is observed in all three periods:
  t = rep(c(-1,0,1), N),
  # Next, generate our group variable such that each observation is in one of two groups:
  G = c(replicate(N, rep(sample(c(0,1),1),3))),
  Z = ifelse(t==1, G, 0),
  # Assign our error term as an idiosyncratic term:
  U = rnorm(N*3), 
  # Generate Y0 as a function of U, G, and t:
  Y0 = 1 + 0.5*t + 1*G + 2*U,
  # Generate Y1 as a function of Y0, the ATE, and t:
  Y1 = Y0 + rnorm(N, ATE, 0.25),
  # Realize Y from PO and Z:
  Y = Y1*Z + Y0*(1-Z)
)

# Let's check our ATT in post-period (given our DGP, = ATE):
mean(data$Y1[data$Z==1]) - mean(data$Y0[data$Z==1])

```

Let's visualise our data and our missing potential outcome:

```{r po_visualise_2x2}
y_sum <- data %>% 
  group_by(t, G) %>% 
  summarise(Y_mean = mean(Y),
            Y_se = sd(Y)/sqrt(sum(!is.na(Y)))) %>%
  ungroup()%>%
  mutate(Outcome = c(rep(c("Y|G=0","Y|G=1"),3)))

y0_sum <- data %>% 
  filter(G == 1) %>%
  group_by(t, G) %>%
  summarise(Y_mean = mean(Y0),
            Y_se = sd(Y0)/sqrt(sum(!is.na(Y0)))) %>%
  ungroup()%>%
  mutate(Outcome = c(rep("Y(0)|G=1", 3)))

y_plot <- y_sum %>%
  rbind(y0_sum)

ggplot(y_plot, aes(x = t, y = Y_mean, colour = Outcome)) + 
  geom_line(alpha = 0.9) + 
  geom_point(alpha = 0.9) + 
  geom_errorbar(aes(ymin = Y_mean - 1.96*Y_se, ymax = Y_mean + 1.96*Y_se), width = 0) +
  labs(title = "Diff-in-Diff", x = "Time", y = "Y", legend = "Treatment Group") +
  theme_minimal()

```

## Analysis

What happens if we just leverage G:

```{r naive_analysis_2x2}
summary(lm_robust(Y ~ G, data))

```

Now let's try a implement the diff-in-diff:

```{r plug_in_estimator_2x2}
means <- data %>%
  group_by(G, t) %>%
  summarise(Y_mean = mean(Y),
            Y_se = sd(Y)/sqrt(sum(!is.na(Y)))) %>%
  ungroup()

# Let's do our diff-in-diff:
(means$Y_mean[means$G==1 & means$t==1] - means$Y_mean[means$G==1 & means$t==0]) - (means$Y_mean[means$G==0 & means$t==1] - means$Y_mean[means$G==0 & means$t==0])

# We can re-arrange the terms:
(means$Y_mean[means$G==1 & means$t==1] - means$Y_mean[means$G==0 & means$t==1]) - (means$Y_mean[means$G==1 & means$t==0] - means$Y_mean[means$G==0 & means$t==0])


```

We can do the same exercise using our regression specifications:

```{r regression_2x2}
# Standard regression estimator
lm_robust(Y ~ G*t, data[data$t > -1,])

# First-differences estimator (point estimate equivalent when t = 2, but notice change in SE):
data_fd <- data %>%
  filter(t > -1) %>%
  group_by(id) %>%
  # Note that lag() is masked, so must call specifically from dplyr:
  mutate(Y_change = Y - dplyr::lag(Y))

lm_robust(Y_change ~ Z, data_fd)

# Two-way fixed effects estimator (LSDV -- same point estimate, but notice (Intercept) has changed -- why?):
lm_robust(Y ~ Z + factor(id) + factor(t), data[data$t > -1,])$coef[1:10]

```

Finally, let's try and falsify the design by checking parallel pre-trends:

```{r parallel_pre_trends_2x2}
# Standard regression estimator
data_pretrend <- data %>%
  filter(t < 1) %>%
  mutate(t = t + 1)

lm_robust(Y ~ G*t, data_pretrend)

# Note, we could just have done this the following way, but it might not work so well in more complex settings:
lm_robust(Y ~ G*t, data[data$t < 1,])

```

# Multi-Period Difference-in-Differences

Simulate data from multiple periods, where treatment is staggered but is absorbing.

```{r simulate_data_multi}
N = 1000
periods = 10

data_multi = tibble(
  id = sort(rep(1:N, periods)),
  t = rep(0:(periods-1), N),
  U = rnorm(N*periods), 
  # Assign a never-treated or pure control group:
  C = c(replicate(N, rep(sample(c(1,0),1, prob = c(0.2,0.8)),periods))),
  # Assign everyone a treatment group cohort (we will remove those who are never treated later):
  G = c(replicate(N, rep(sample(c(0:periods),1),periods))),
  # Assign treatment to those who are treated in their post-period:
  Z = ifelse(t >= G, 1, 0),
  # Assign untreated potential outcome
  Y0 = 1 + 0.5*t + 1*G + 2*U,
  # Assign the effect, but do so with heterogeneity over time:
  Y1 = case_when(
    t < 5 ~ Y0 + (rnorm(N*periods, 2, 0.25)*(t^(0.75))),
    t >= 5 & t < 8 ~ Y0 + (rnorm(N*periods, 2, 0.25)*(t^(0.5))),
    t >= 8 ~ Y0 + (rnorm(N*periods, 2, 0.25)*(t^(0.1))),
  ),
  Y = Y1*Z + Y0*(1-Z)
  ) %>%
  mutate(G = ifelse(C == 0, G, 0),
         Z = ifelse(C == 0, Z, 0))

# visualise the treatment regime (staggered and absorbing) and the outcomes:
panelview(data_multi, index = c("id","t"), Y = "Y", D = "Z", main = "Panel View of Data", xlab = "Time", ylab = "Units", 
          display.all = TRUE)
panelview(data_multi, index = c("id","t"), Y = "Y", D = "Z", main = "Panel View of Data", xlab = "Time", ylab = "Units", 
          type = "outcome")

# Check our estimand, the ATT in the post-period: 
mean(data_multi$Y1[data_multi$Z==1]) - mean(data_multi$Y0[data_multi$Z==1])

# Visualize heterogeneous effects over time:
ypo_sum <- data_multi %>%
  group_by(t) %>%
  summarise(hetfx = mean(Y1 - Y0))

ggplot(ypo_sum, aes(x = t, y = hetfx)) +
  geom_point() +
  theme_minimal()

```

```{r analysis_multi}
# Start with a pooled OLS model:
lm_robust(Y ~ Z, data_multi)

# What if we add unit fixed effects (slow!):
lm_robust(Y ~ Z + factor(id), data_multi)$coef[1:10]

# Let's try two-way fixed effects instead:
# First, we can use the fixest package, which is much faster:
twfe_feols <- fixest::feols(Y ~ Z | id + t, data_multi, cluster = "id")
summary(twfe_feols)

# Or we can use PLM (you would need to adjust the SEs for clustering)
twfe_plm <- plm::plm(Y ~ Z, data = data_multi, index = c("id","t"), model = "within", effect = "twoways")
summary(twfe_plm)

```

Finally, let's try a 'modern' estimator, from the fect package:

```{r fect}
# Use fect package for fast estimation of fixed effects (you should increase the nboots)
fect_out <- fect::fect(Y ~ Z, data = data_multi, index = c("id","t"), method = "ife", se = TRUE, nboots = 50)

# Visualise the effect:
plot(fect_out)

# Test for pre-trends 1 -- joint F-test:
plot(fect_out, type = "equiv")

# Placebo period tests:
fect_out_p <- fect::fect(Y ~ Z, data = data_multi, index = c("id","t"), method = "ife", se = TRUE, nboots = 50, placebo = TRUE, placebo.period = c(-2,0))

plot(fect_out_p, stats = c("placebo.p","equiv.p"))

```

Finally, let's try with Callaway & Sant'Anna's estimator, using the did package:

```{r csdid}
# Pre-process the data 
did_proc_data <- data_multi %>%
  did::pre_process_did(yname = "Y", tname = "t", idname = "id", gname = "G", allow_unbalanced_panel = T, data = .)

# Estimation using pre-processed data:
# Note, you should use base_period = 'universal' even though this is not the default. See Roth, 2024, "Interpreting Event-Studies from Recent Difference-in-Differences Methods"
reg <- did_proc_data[['data']] %>%
  did::att_gt(yname = "Y", tname = "t", idname = "id", gname = "G", 
         control_group = c("nevertreated"), est_method = 'dr', base_period = 'universal',
         allow_unbalanced_panel = T, data = .)

reg

ggdid(reg,ncol=3)

```