---
title: "MY457/MY557: Causal Inference for Experimental and Observational Studies"
subtitle: "Class 4: Instrumental Variables"
author: ""
date: ''
output:
  html_document: default
  pdf_document: default
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,warning=F,message=F}
library(dplyr)
library(ggplot2)
library(AER)
```

# In-class exercise: Implementing intrumental variables estimation

In this exercise we try different ways of calculating the basic instrumental variables estimator. This is mainly for demonstration purposes. In practice, it is best to use the `ivreg` function, because it also implements the standard errors correctly. At first, let's start with creating a dataset.

```{r}
# SIMULATE DATA

# PARAMETERS
N <- 10000
U <- rnorm(N, mean = 5, sd = 3)
b0 <- 2
b1 <- 1.5

# POTENTIAL OUTCOMES
y0 <- b0 + b1 * U + rnorm(N)
y1 <- y0 + mean(y0) + rnorm(N)
y1[which(y0 < median(y0))] <- y1[which(y0 < median(y0))] / 2

# CREATE DATAFRAME
df <- cbind(y0, y1, U) %>% as_tibble()

# GENERATE TYPES OF COMPLIANCE
type <- rep(NA, 10000)
type[which(y1 > median(y1))] <-
  sample(c(rep('Complier', 3500), rep('Always Taker', 1000), rep('Never Taker', 500)))
type[which(y1 < median(y1))] <-
  sample(c(rep('Complier', 1500), rep('Always Taker', 1500), rep('Never Taker', 2000)))
df$type <- type

# CREATE INSTRUMENT
df$z <- sample(c(rep(0, 5000), rep(1, 5000)))

# CREATE TREATMENT ASSIGNMENT
df <- df %>%
  mutate(d = case_when(type == 'Always Taker' ~ 1,
                       type == 'Never Taker' ~ 0,
                       (type == 'Complier' & z == 0) ~ 0,
                       (type == 'Complier' & z == 1) ~ 1))

# REAL OUTCOMES
df <- df %>% mutate(y = case_when(d == 0 ~ y0, d == 1 ~ y1))
```


Before we estimate the treatment effect via the instrumental variable Z, let's use the potential outcomes to calculate the average treatment effects.
```{r}
# TRUE ATE
true_ate <- t.test(df$y1, df$y0, paired = TRUE)
true_ate
```

In contrast, the naive approach would be if we just regress the observed outcome on the treatment assignment indicator.
```{r}
# NAIVE "ATE"
naive_ate <- lm(y ~ d, data = df)

true_ate
summary(naive_ate)
```


Now, because we do not observe both potential outcomes, we need a different strategy. In the following, we will the instrumental variable approach to get a local average treatment effect (LATE). There are basically three different estimates we need to understand: (i) the effect of Z on Y (the intention-to-treat, ITT), (ii)  the effect of Z on D (estimated proportion of compliers), (iii) the Wald estimate.
```{r}
# 1. Effect of Z on Y
mean(df$y[df$z==1])-mean(df$y[df$z==0])
y.on.z <- lm(y ~ z, data = df)
summary(y.on.z)
itt_est <- coef(y.on.z)[2]

###

# 2. Effect of Z on D
mean(df$d[df$z==1])-mean(df$d[df$z==0])
d.on.z <- lm(d ~ z, data = df)
summary(d.on.z)
prop_compliers <- coef(d.on.z)[2]

###

# 3. WALD ESTIMATE
itt_est/prop_compliers
```

However, you can also get the same number from the two-stage least squares (2SLS), (i) with the base-R lm function, or (ii) with the ivreg function.

```{r}
# 2SLS using lm
df$d_hat <- predict(d.on.z)
iv_2sls_2 <- lm(y ~ d_hat, data = df)
summary(iv_2sls_2)

# 2SLS using ivreg
iv_2sls <- ivreg(y ~ d | z, data = df)
summary(iv_2sls)
```

Looking at the data structure above, it is important to note that in a real-world application, one would only observe $Z$, $D$, and $Y$. We are only able to observe $Y_{0}$, $Y_{1}$, $U$, and $Type$ because this is a simulation exercise and we generated them ourselves.

Further, we can briefly examine the distribution of the true individual-level treatment effects according to type and treatment status.
```{r}
df <- df %>% mutate(y_diff = y1 - y0)
p1 <- ggplot(data = df, aes(x = y_diff, fill = type)) +
  geom_density(alpha = 0.2) + ggtitle("All Units")
p1
p2 <- ggplot(data = df[df$d == 1, ], aes(x = y_diff, fill = type)) +
  geom_density(alpha = 0.2) + ggtitle("Treated Units")
p2
p3 <- ggplot(data = df[df$d == 0, ], aes(x = y_diff, fill = type)) +
  geom_density(alpha = 0.2) + ggtitle("Control Units")
p3
```

The plots indicate that being a complier is associated with taking on relatively high values of the $Y_{1}$ potential outcome and therefore relatively high values of the true treatment effect. The bimodal distributions indicate that substantial proportions of each type take on both relatively low and relatively high values of the treatment effect, but clearly one side or the other dominates.

